Program Transformations for miniKanren

### 2020-07-05 papers/aplas/text/evaluation.tex
Maxlen subsection

The relation \lstinline{maxLenght$^o$} computes both the length of the list and its maximum element (see fig.~\ref{cpd:maxandlength}) by concatenation of two relation calls.
This relation traverses the list \lstinline{xs} twice: first to compute the maximum and then to compute the length of the list.
It can be transformed in such a way that the list is only traversed once while computing both components of the result simultaneously.
This transformation is called \emph{tupling} and can be achieved with conjunctive partial deduction.

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.85\textwidth}
\begin{lstlisting}[label={cpd:maxandlength}, caption={Maximum element and length of the list}, captionpos=b, frame=tb]
  let maxLength$^o$ xs m l = max$^o$ xs m /\ length$^o$ xs l
  let rec length$^o$ xs l = conde [
    (xs === [] /\ l === zero);
    (fresh (h t m) (
      xs === h % t /\ l === succ m /\ length$^o$ t m))]
  let max$^o$ xs m = max$_1^o$ xs zero m
  let rec max$_1^o$ xs n m = conde [
    (xs === [] /\ m === n);
    (fresh (h t) (
      (xs === h % t) /\
      (conde [
        (le$^o$ h n ^true /\ max$_1^o$ t n m);
        (gt$^o$ h n ^true /\ max$_1^o$ t h m)])))]
  let rec le$^o$ x y b = conde [
    (x === zero /\ b === ^true);
    (fresh (x$_1$) (x === succ x$_1$ /\ y === zero /\ b === ^false));
    (fresh (x$_1$ y$_1$) (x === succ x$_1$ /\ y === succ y$_1$ /\ le$^o$ x$_1$ y$_1$ b))]
  let rec gt$^o$ x y b = conde [
    (x === zero /\ b === ^false);
    (fresh (x$_1$) (x === succ x$_1$ /\ y === zero /\ b === ^false));
    (fresh (x$_1$ y$_1$) (x === succ x$_1$ /\ y === succ y$_1$ /\ gt$^o$ x$_1$ y$_1$ b))]
  \end{lstlisting}
\end{minipage}
\end{figure*}

The ideal implementation is shown in fig.~\ref{ideal:maxandlength}.
It uses a recursive relation to compute the length and maximum element simultaneously while traversing the list once.

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.7\textwidth}
\begin{lstlisting}[label={ideal:maxandlength}, caption={Ideal implementation of maxlengtho}, captionpos=b, frame=tb]
  let maxLength$^o$ xs m l = maxLength$_1^o$ xs m zero l
  let rec maxLength$_1^o$ xs m n l = conde [
    (xs === [] /\ m === n /\ l === zero);
    (fresh (h t l$_1$)
       (xs === h % t) /\
       (l === succ l$_1$) /\
       (conde [
         (le$^o$ h n ^true /\ maxLength$_1^o$ t m n l);
         (gt$^o$ h n ^true /\ maxLength$_1^o$ t m h l)]))]
  \end{lstlisting}
\end{minipage}
\end{figure*}

We measured the execution time of \lstinline{maxLength$^o$ lst m l}, where \lstinline{lst} is a list of Peano numbers from $1$ to $100$.
Note, that the second disjunct in the implementation of both \lstinline{le$^o$} and \lstinline{gt$^o$} can never contribute into the result of \lstinline{maxLength$^o$} execution.
Thus among others we compared two ideal implentations: one includes these disjuncts, while the other removes them (see table~\ref{tbl:maxlen})

\begin{table}
  \centering
  \begin{tabular}{c||c}
                   & [1..100] \\ \hline\hline
  Original         & 2.286ms  \\ \hline
  Ideal           & 4.623ms  \\ \hline
  Ideal removed   & 3.133ms  \\ \hline
  \ecce             & 1.701ms  \\ \hline
  ConsCPD          & 2.578ms
  \end{tabular}

@@

Surprising enough, the execution time of the ideal program is worse than of the original.
\ecce succeeds at improving the program performance, while conservative partial deduction worsens it a little.
\todo{why}

>>>>

The relation \lstinline{maxLenght$^o$} computes the maximum element of the input list and its length (see fig.~\ref{cpd:maxandlength}) by conjunction of two relation calls.
The list \lstinline{xs} is traversed twice: first to compute the maximum in the relation \lstinline{max$^o$} and then to compute the length in the relation \lstinline{length$^o$}.
The input list contain Peano numbers which are compared by relations less-or-equal (\lstinline{le$^o$}) and greater-than (\lstinline{gt$^o$}).
Note that the comparison relations have three arguments with the last one representing the result of comparison.

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.85\textwidth}
\begin{lstlisting}[label={cpd:maxandlength}, caption={Maximum element and length of the list}, captionpos=b, frame=tb]
  let maxLength$^o$ xs m l = max$^o$ xs m /\ length$^o$ xs l
  let rec length$^o$ xs l =
    fresh (h t m) ( conde [
      (xs === [] /\ l === zero);
      (xs === h % t /\ l === succ m /\ length$^o$ t m)])

  let max$^o$ xs m = max$_1^o$ xs zero m
  let rec max$_1^o$ xs n m =
    fresh (h t) ( conde [
      (xs === [] /\ m === n);
      (xs === h % t) /\
      (conde [
        (le$^o$ h n ^true /\ max$_1^o$ t n m);
        (gt$^o$ h n ^true /\ max$_1^o$ t h m)])])

  let rec le$^o$ x y b =
    fresh (x$_1$ y$_1$) ( conde [
      (x === zero /\ b === ^true);
      (x === succ x$_1$ /\ y === zero /\ b === ^false);
      (x === succ x$_1$ /\ y === succ y$_1$ /\ le$^o$ x$_1$ y$_1$ b)])
  let rec gt$^o$ x y b =
    fresh (x$_1$ y$_1$) ( conde [
      (x === zero /\ b === ^false);
      (x === succ x$_1$ /\ y === zero /\ b === ^true);
      (x === succ x$_1$ /\ y === succ y$_1$ /\ gt$^o$ x$_1$ y$_1$ b)])
  \end{lstlisting}
\end{minipage}
\end{figure*}

The ideal implementation provided in the CPD literature~\cite{de1999conjunctive} is shown in fig.~\ref{ideal:maxandlength}.
This implementation traverses the input list \lstinline{xs} only once and can be generated by tupling.
Note, that the disjunct which unifies the third argument with \lstinline{^false} in the implementation of both \lstinline{le$^o$} and \lstinline{gt$^o$} can never contribute into the result of \lstinline{maxLength$^o$} execution.
Such disjuncts are removed by both \ecce and conservative partial deduction, thus we consider two ideal implementions: one with the original comparison relations (\emph{Ideal}) and another in which these disjuncts are deleted (\emph{Ideal removed}).

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.7\textwidth}
\begin{lstlisting}[label={ideal:maxandlength}, caption={Ideal implementation of maxlengtho}, captionpos=b, frame=tb]
  let maxLength$^o$ xs m l = maxLength$_1^o$ xs m zero l
  let rec maxLength$_1^o$ xs m n l =
    fresh (h t l$_1$) ( conde [
      (xs === [] /\ m === n /\ l === zero);
      ((xs === h % t) /\
       (l === succ l$_1$) /\
       (conde [
         (le$^o$ h n ^true /\ maxLength$_1^o$ t m n l);
         (gt$^o$ h n ^true /\ maxLength$_1^o$ t m h l)]))])
  \end{lstlisting}
\end{minipage}
\end{figure*}

We measured the execution time of \lstinline{maxLength$^o$ lst m l}, where \lstinline{lst} is a list of Peano numbers from $1$ to $100$.
The results are presented in table~\ref{tbl:maxlen}.

\begin{table}
  \centering
  \begin{tabular}{c||c}
                   & [1..100] \\ \hline\hline
  Original         & 48.5ms  \\ \hline
  Ideal            & 32.2ms  \\ \hline
  Ideal removed    & 19.6ms  \\ \hline
  \ecce            & 23.7ms  \\ \hline
  ConsCPD          & 35.7ms
  \end{tabular}

@@

\includegraphics{graphs/maxlen.pdf}

Both \ecce and ConsPD improve performance of \lstinline{maxLength$^o$} as compared to the original program.
Removing the disjuncts with the result unified with \lstinline{^false} in the comparison relations improves the ideal program by about 40\%.
ConsPD does not perform tupling, but it does remove disjuncts which never succeed, thus its result is not as performant as the ideal program.
